---
title: "Objects & Classes"
subtitle: "Part 1"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = F,
                      message = F,
                      fit.retina = 3,
                      fig.align = "center")

hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#4834d4")
```

```{r starwarsData, include=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

empire <- starwars

empire <- empire[c(1:5, 10, 13, 14, 19, 21), c(1:3, 8:10)]
empire$species <- factor(empire$species)

```

# Plan for today

- What is an **object**?

--

- **Assignment**

--

- More than 1 thing with **vectors**

--

- Object **classes**

---

# What do we *want*?

We *want* our data to look something like this...

```{r swDataEx, echo=FALSE}
kable(empire, format = "html")
```

---

# What do we *want*?

What R sees...

```{r swDataEx2, echo=FALSE}
empire
```

---
name: objects

# How do we get there? 

### Objects

- A basic concept in (statistical) programming is called an **object**

- An **object** allows you to store a value or a thing:

---

# An object can be...

```{r swObjects, echo=FALSE}

empire$mass[4] <- cell_spec(empire$mass[4], "html", background = c("#FFFF00"))

kable(empire, format = "html", escape = F)

```

---
# An object can be...

```{r swObjects2, echo=FALSE}

empire$mass[4] <- "136"

kable(empire, format = "html", escape = F) %>%
  column_spec(3, background = c("#FFFF00"))

```

---

# An object can be...

```{r swObjects3, echo=FALSE}

kable(empire, format = "html", escape = F) %>%
  row_spec(row = 4, background = c("#FFFF00"))

```

---

# An object can be...

```{r swObjects4, echo=FALSE}

kable(empire, format = "html", escape = F) %>%
  row_spec(row = c(1:10), background = c("#FFFF00"))

```

---

### <span style="color: #be2edd;">Important</span>:

- Objects have *names*
- We are going to refer to objects by their *names*
- Since they have names, we can **store** objects and use them later

---
name: assignment

# Storing data in objects

If you want to use an object later on (you do!), you have to name it.

This is called **assignment** or **assigning** a name to an object

In takes the form of:

```{r assign, eval=FALSE}

nameOfMyObject <- objectToStore

```

---
class: inverse, middle, center
name: practice1

# Go do Practice #1

When you have finished, come back here.

---

# What's the point of storing objects?

Remembering things sucks! Let R hold on to all the stuff you don't want to remember or write down right away. 

Let's do an example with a series of math equations:
$$y = 17*8$$
$$z = \frac{y}{3}$$
How do we solve this?:

1. Solve for $y$, which is $136$. Either remember $136$ or write the number down.
2. Plug it in in the second equation, so that you have $136$ divided by $3$. 

---

#Let's do this with R code!

Let's do an example with a series of math equations:
$$y = 17*8$$
$$z = \frac{y}{3}$$

With code:
```{r eq1}
y <- 17*8 # first, solve for y 
z <- y/3 # now, solve for z 
```

--

We didn't even need to know that `17*8` is `136`. We stored the value of `136` as an object with the name `y`.

--

Then, we could tell R to simply use the name `y` anytime we wanted to refer to the number `136` 

---
# Who cares?

Remembering a single number seems a little ridiculous. But remember, an object in R can really be anything. Some objects you definitely might want to store for later:

- A data set like `empire`
- A correlation coefficient
- The output of a linear regression model
- $p$-values and other statistics
- The mean of a variable, so you can subtract the mean from every individual's score
- and lots, lots more!
---

class: center

## If you do not assign a name to an object, R will not remember it!

Example:
```{r, include=FALSE}
rm(y,z)
```

```{r, error=TRUE}
17*8

y/3
```

--
<br>
<br>

#### The error message `object 'y' not found` is very common!
R cannot perform the operation because you never told it to remember `17*8`
---

class: inverse, middle, center
name: practice2

# Go do Practice #2

When you have finished, come back here.

---
name: vector

# One type of object: **Vectors**

A group of objects is called a **vector**

**Vectors** are *ONE-DIMENSIONAL*. You can think of this as either a row...

```{r vec1, echo=FALSE}

kable(empire, format = "html", escape = F) %>%
  row_spec(row = 4, background = c("#FFFF00"))

```

---

# One type of object: **Vectors**

A group of objects is called a **vector**

**Vectors** are *ONE-DIMENSIONAL*. You can think of this as either a row...
... or a column

```{r vec2, echo=FALSE}
kable(empire, format = "html", escape = F) %>%
  column_spec(3, background = c("#FFFF00"))

```

---
# Making Vectors
In your R code, you will type `c()` in order to create a vector

The `c` stands for *"combine"* or *"concatenate"*

Some examples:
```{r vec3}
subjectID <- c("Subject 1", "Subject 2", "Subject 3", "Subject 4", "Subject 5")
passedStats <- c(TRUE, FALSE, FALSE, TRUE, TRUE)
favoriteNumbers <- c(7, 3, 6, 10, 100)
countries <- c(0, 3, 10, 1, 8)

```

---

# Vectors
Because these items are grouped together, you can do something to them all at once!

Let's say these 5 people all went on a trip together, and they visited 2 countries. We can add `2` to the entire vector, rather than each individual number:

```{r vec4}
countries + 2
```

---

class: inverse, middle, center
name: practice3

# Go do Practice #3

When you have finished, come back here.

---

name: classes
# Basic object classes

Objects can be of a different `class`. You can think of it more as *what type of information is stored in the object?*. Some of the options are:

- **Numeric:** Decimals (3.141593)

- **Integer:** Natural numbers (0,1,2, etc.)

- **Character:** Text or string characters:

    - Always inside quotation marks
    - **Factors** (or categories)
- **Logical:** True or False:

    - No quotations
    - 2 possible values: `TRUE` or `FALSE`
    
- **Missing Value:** `NA`

