---
title: "Indexing"
subtitle: "Part 1"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer-yellow.css", "css/additionalCols.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = F,
                      message = F,
                      fit.retina = 3,
                      fig.align = "center")

hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#f9ca24",
                  outfile = "xaringan-themer-yellow.css")
```

```{r starwarsData, include=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

empire <- starwars

empire <- empire[c(1:5, 10, 13, 14, 19, 21), c(1:3, 8, 10, 11)]
empire$species <- factor(empire$species)

```

# Last time was .color-light-3[objects]

- **.color-light-3[Objects]** are the **.color-light-3[nouns]** of programming languages

--

- **.color-light-3[Objects]** store something
  - value
  - character string
  - **.color-light-3[vector]** of values or characters
  - entire data frame
  - model
  - etc...

---
# So what? 
- Making an .color-light-3[object] is only useful if you can use it later on
--

- We need to be able to access our .color-light-3[objects]
--

- And we need to be able to access bits and pieces of our .color-light-3[objects]
  - Example: what if you only want one or a few of the objects in a **.color-light-3[vector]**? 

---

name: indexing

# Indexing

#### **Indexing** is how we access items within a .color-light-3[vector].
--

To index a **.color-light-3[vector]**, we're going to use square brackets `[ ]`
---
# An example

Let's make a **.color-light-3[vector]** of *some* my favorite TV `comedies`:
```{r comedies0}
comedies <- c("Parks and Rec", "Broad City", "I'm Sorry", "Avenue 5",
              "Bored to Death", "The Office", "Silicon Valley", "30 Rock",
              "Big Mouth", "Futurama")
```

--
If you want to look at *all* the objects contained in the vector, all you need is your vector name
```{r comedies1}
comedies
```
--

But, if I only want to know the 6th TV show in this `comedies` vector I would type:
```{r comedies2}
comedies[6]
```

---

# Indexing

An **index** is the *position* of an element in a .color-light-4[vector]

--

In `R`, indexing starts at 1. That is, the first element of a vector is located at position 1. 

--

The basic syntax is to take the name of the .color-light-4[vector] and directly next to in, in square brackets, put the position you want to get. 

  - Something like `objectName[position number]`

---
name: indexingVectors

# Indexing Vectors

```{r comedies3}
comedies <- c("Parks and Rec", "Broad City", "I'm Sorry", "Avenue 5",
              "Bored to Death", "The Office", "Silicon Valley", "30 Rock",
              "Big Mouth", "Futurama")
```

To select objects that are sequential (in a row):

  - `comedies[3:5]`
  - You can think of `:` as *"through"*
    - `[3:5]` = "three *through* five"
--

```{r comedies4}
comedies[3:5]
```

---

# Indexing Vectores

To select objects that are *not* in a row:
  
```{r comedies5}
comedies[c(1,7,10)]
```
--

Note that you need to wrap the position numbers inside of `c()`. What you're doing is making a mini-vector of position numbers! Vectors on vectors!
---

# Indexing Vectors
Since you can make these mini-vectors, you can go crazy and combine sequential and non-sequential objects. 
  - Example:
  
```{r comedies6}
comedies[c(1:4, 9)]
```

This reads as *"give me the elements that correspond to positions 1,2,3,4 (1 through 4), and 9"*

---
name: impact

# A more serious example
Let's say we have a vector that contains body mass indices (BMIs) for 20 participants in your study. A healthy BMI is between 18.5-24.9, and anything above 30 is considered "obese". You inspect your vector called `bmi`.

```{r bmi1, echo=FALSE}
bmiNormal = runif(15, min=18, max=35)
bmiOutrageous = runif(5, min=50, max=100)

bmi = round(c(bmiNormal, bmiOutrageous), digits = 1)

bmi
```

--

The last 5 numbers are WAY off the scale! You recently implemented a new survey tool to collect the data, and only those 5 participants used the new method. You realize there's probably something wrong with the data collection!

You'll need those **index** positions in order to remove those 5 participants. 

```{r bmi2}
bmi[16:20]
```

(We will talk more about removing participants in the coming weeks!)

  