---
title: "Basic Statistics & Plotting"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer-orange.css", "css/additionalCols.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = F,
                      message = F,
                      fit.retina = 3,
                      fig.align = "center")

hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#f0932b",
                  outfile = "xaringan-themer-orange.css")
```

```{r starwarsData, include=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

empire <- starwars

empire <- empire[c(1:5, 10, 13, 14, 19, 21), c(1:3, 8:10)]
empire$species <- factor(empire$species)

```

name: recap

# You've already learned *a lot*!
- Objects
- Classes
- Indexing 1-d vectors and 2-d data.frames
- Logical operators & functions
- How to look at the help documentation
- Installing/loading packages
- Making new script files
- Importing your data etc.
- Basic statistics
- Basics of `ggplot2`
---

# This time
- Lots of practice sets; letting it all come together
  
- Some random tidbits to make your life easier
---

# About the MIDUS dataset
Project Description:
  "The purpose of the study was to investigate the role of behavioral, psychological, and social factors in understanding age-related differences in physical and mental health."
--

Variables available in this data file:
  - **Demographic variables**: age, sex
  - **Physical health variables**: self-rated physical health, heart problems, father had heart attack, BMI
  - **Mental health variables**: self-rated meantal health, self-esteem, life satisfaction *(life overall, work, health, relationship with spouse/partner, relationship with children)*, hostility *(stress reactivity & agression*)
---

name: stats

# Statistical Analyses
You already know everything you need to run statistical analyses! All statistics in `R` are computed by using **functions**. If you know the type of analysis you want to run, find the corresponding function and go for it!

*Is there a difference in `hostility` between men and women in this sample?*

.box-inv-2[Statistic: T-test]
---
# A note about 

