---
title: "Visualizing Data"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer-purple.css", "css/additionalCols.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = F,
                      message = F,
                      fit.retina = 3,
                      fig.align = "center")

hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```

```{r, echo = F, warning = F, message = F, results = 'hide'}
library(psych)
library(tidyverse)
library(broom)
```

# Recap

- We have spent **a lot** of time working with `ggplot2`

- Everything covered this week can be applied with `ggplot2`

- Goal of this week is to get you to stop and think about your plots *before* you make them. What are the kinds of things you should be considering? 

---
# Stuff you already know

Data visualization is both an art *and* a science

Art:
  - Color choices
  - The story you're trying to tell
  
Science:
  - Graphical representation of specific types of data
  - 

---
class: inverse, center

# Warning

### Making figures for academic purposes $\neq$ making figures for pure data viz purposes

---
# Academic Figures

Things you need to think about:

- 