---
title: "Advanced Statistical Methods"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer-blue.css", "css/additionalCols.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      navigation:
        scroll: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = F,
                      message = F,
                      fit.retina = 3,
                      fig.align = "center")

hook_source <- knitr::knit_hooks$get('source')
knitr::knit_hooks$set(source = function(x, options) {
  x <- stringr::str_replace(x, "^[[:blank:]]?([^*].+?)[[:blank:]]*#<<[[:blank:]]*$", "*\\1")
  hook_source(x, options)
})
```

```{r, echo = F, warning = F, message = F, results = 'hide'}
library(psych)
library(tidyverse)
library(broom)
library(knitr)
```

## Recap

- If you had to go back to Psych 300, you should be able to do everything in `R`
- You should feel better about the stuff you learned in Psych 300
- You maybe learned something new, especially about interactions in regression

--

## This Time

- Where your research might lead you
- This is a fun lecture. This is not to scare you.
- I will *NOT* show you how to do everything in `R`, but I will provide resources
- Sit back and enjoy the ride

---
name: mlm

## Multi-level Modeling

AKA:

- linear mixed effects models
- hierarchical linear models

--

**We use this when our data are _nested_**

