---
title: "Descriptives - Redux"
linktitle: "Descriptives"
menu:
  lectures:
    parent: Statistics
    weight: 1
type: docs
slides: "20-slides"
output:
  blogdown::html_page:
    toc: true
---

## Videos

- [Descriptive Statistics Redux](https://youtu.be/zceFiIEsU8E)
- In the last 10 minutes, my video froze `r emo::ji("sad")` At that point, I remove the video feed, but the audio still works normally. So you can expect my face to disappear lol! 

## Data

This lecture will use a world happiness dataset. If you want to play along, you can download it here:

- [<i class="fas fa-file-csv"></i> `world_happiness_2015.csv`](/slides/20-slides_files/world_happiness_2015.csv)


## Slides

`r blogdown::shortcode("slide-buttons")`

```{r show-slide-tabs, echo=FALSE, results='asis'}

source(here::here("R", "slide_tabs.R"))

sections <- tibble::tribble(
  ~name, ~slide, ~active,
  "Next 5 Weeks", "1", TRUE,
  "Central Tendency", "centTen", FALSE,
  "Dispersion", "disp", FALSE,
  "Bias", "bias", FALSE,
  "Standardization", "zs", FALSE
)
slide_tabs(sections, "/slides/20-slides.html")
```

## Extra Slides for Week 8

`r blogdown::shortcode("slide-buttons")`

```{r show-slide-tabs2, echo=FALSE, results='asis'}

source(here::here("R", "slide_tabs.R"))

sections <- tibble::tribble(
  ~name, ~slide, ~active,
  "Validity", "1", TRUE
)
slide_tabs(sections, "/slides/ExtraSlides-Week8.html")
```