---
title: "Practice Set: Objects"
linktitle: "2: Objects"
output:
  blogdown::html_page:
    toc: true
menu:
  practice:
    parent: The Basics
    weight: 2
type: docs
weight: 2
editor_options: 
  chunk_output_type: console
shiny: true
---
```{r setup, warning = FALSE, message = FALSE, include=FALSE}
library(tidyverse)
library(flair)
library(knitr)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4.5, fig.align = "center")
```

## Visualizing Star Wars

Now that you're oriented to RStudio, it's time to write and execute some code. The following uses a data set about Star Wars characters (yay!). This is the same dataset as the [Objects lecture](/lectures/02-lecture)

:::puzzle
**You try:**

  1. Read through the lines of code. You probably won't understand anything, but that's OK!
  2. Click "Run Code" 
  3. Watch what happens!
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/learnr_app1/",
                    id = "learnr-02-starwars")

```

Over the course of this basic training, we'll cover all the elements in this code block that produced this graph. It's 100% OK if it looks like complete gibberish to you right now!

## Practice with Objects
- **Objects** allow you to store things (values, models, entire datasets etc.)
- They are the *nouns* of programming languages; they are typically what we are acting *on*
- We give an object a name; this is called **assignment**, and it takes the form of:

```{r obj-form, eval=FALSE, include=FALSE}
nameOfMyObject <- objectToStore
```

```{r decorate-obj-form, echo=FALSE}
decorate("obj-form", eval=FALSE) %>% 
  flair("nameOfMyObject", background = "yellow") %>% 
  flair("<-", background = "red") %>% 
  flair("objectToStore", background = "cornflowerblue") %>% 
  knit_print.with_flair()
```

The thing on the LEFT (yellow) is the name you want to give your object.

Then the arrow (pink); first `<`, then `-` with no space in between.

The thing you want to actually store for later goes on the RIGHT (blue). 

:::puzzle
**You try!** Create an object called `continents` that stores the number `7`
:::

:::note
**Note:**

  - Make sure you hit "Run Code" *before* "Submit Answer"
  - If you get a message that's something like "Last value is invisible", you can totally 100% ignore it
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/objectsPart1_app2/",
                    id = "learnr-02-objects")

```

## Practice with Vectors
A **vector** is a group of objects that's 1-dimensional:

  - It can be a row, a column, or just stand alone 
  - To make a vector, you can *combine* or *concatenate* objects together using `c()`
  - Each object (vector or otherwise) has a specific data class. this can be things like numeric, character, factors, logical etc.

Check out this table:

```{r, echo=FALSE}
ids <- c("ID1", "ID2", "ID3", "ID4", "ID5")
state <- c("Missouri", "Iowa", "Missouri", "Idaho", "Maine")
smoker <- c(TRUE, FALSE, FALSE, TRUE, FALSE)
age <- c(20, 18, 32, 25, 25)

testdf = data.frame(ID = ids,
                    State = state,
                    Smoker = smoker,
                    Age = age)

kable(testdf)
```

Here, each row can be considered it's own vector, and each column can be considered it's own vector. 

:::puzzle
**You try:**

  1. First, create 4 vectors for each of these columns. Make sure they contain the same exact values as what you see in this table! Assign each vector to the following corresponding name:
   - `ids`
   - `state`
   - `smoker`
   - `age`
  2. Then, check the `class()` of each of your newly created vectors.
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/vectors_app3/",
                    id = "learnr-03-vectors")

```

#### Remember Star Wars?
You've already made a factor! You turned `species` into a factor. The result is that in our graph, we have discrete categories for the different types of Star Wars characters!

```{r empire-code, include=FALSE}
empire <- starwars

empire <- empire[c(1:5, 10, 13, 14, 19, 21), c(1:3, 8, 10, 11)]
empire$species <- factor(empire$species)

ggplot(data = empire, aes(x = mass, y = height)) +
  geom_point(aes(size = mass,
                 color = species),
             alpha = .5) +
  labs(title = "Star Wars Characters",
       subtitle = "by size",
       x = "Weight (kg)",
       y = "Height (cm)") +
  scale_size(guide = "none")
```

```{r decorate-empire-code, echo=FALSE}
decorate("empire-code") %>% 
  flair("empire$species <- factor(empire$species)",
        background = "yellow") %>% 
  flair("color = species") %>% 
  knit_print.with_flair()
```

