---
title: "Practice Set: Indexing"
linktitle: "3: Indexing"
output:
  blogdown::html_page:
    toc: true
menu:
  practice:
    parent: The Basics
    weight: 3
type: docs
weight: 3
editor_options: 
  chunk_output_type: console
shiny: true
---
```{r setup, warning = FALSE, message = FALSE, include=FALSE}
library(tidyverse)
library(flair)
library(knitr)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4.5, fig.align = "center")
```

## Indexing Recap

In the [previous practice set](/practice/02-practice), you created the following 4 vectors:   

```{r}
ids <- c("ID1", "ID2", "ID3", "ID4", "ID5") # character
state <- c("Missouri", "Iowa", "Missouri", "Idaho", "Maine") # character/factor
smoker <- c(TRUE, FALSE, FALSE, TRUE, FALSE) # logical
age <- c(20, 18, 32, 25, 25) # numeric
```

You can do something to *all* elements of the vector:

```{r}
# add 10 years to everyone's age
age <- age + 10
age
```

If you have 2 vectors, these can be combined in a number of ways!

Let's try to add the number `2` to the first person's age, `4`, to the second person's age etc.
```{r}
numbers <- c(2, 4, 6, 8, 10)

age + numbers
```

You can also append one vector on to the next. Notice that to do this, we use `c()`. We're `concatenating` or `combining` the two vectors.  
```{r}
newVector <- c(age, numbers)
newVector
```

If you have 2 vectors, but they are not the same length, R will recycle the shorter vector.
```{r}
shortVector <- c(1000, 2000)

newVector + shortVector
```

If the longer vector is not a multiple of the shorter vector, the process still works but you'll also get a warning message (notice how it still works, though):

```{r, error=FALSE}
age + shortVector
```

:::puzzle
**You try!** The following exercises are a modified version of the `sat.act` dataset (but broken up into vectors). The SAT & ACT are both standardized college admissions tests in the United States. The ACT technically has 4 parts, but only the overall score is reported here. The SAT is broken up into verbal and quantitative categories:

  - `age` in years
  - `act` contains scores from the ACT tests; range 1-36, norm ~20
  - `satv` contains the Verbal scores from the SAT; range 200-800, avg ~500
  - `satq` contains the quantitative scores from the SAT; range 200-800, avg ~500

Follow the directions above each code chunk!
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/03-learnr-app1/",
                    id = "learnr-03-app1")

```



