---
title: "Practice Set: Interactions"
linktitle: "Interactions"
output:
  blogdown::html_page:
    toc: true
    css: "/slides/css/additionalCols.css"
menu:
  practice:
    parent: Statistics
    weight: 6
type: docs
weight: 6
editor_options: 
  chunk_output_type: console
shiny: true
---

```{r setup, warning = FALSE, message = FALSE, include=FALSE}
library(tidyverse)
library(flair)
library(knitr)
library(kableExtra)
library(datasets)
library(sjPlot)
library(reghelper)
library(ggeffects)
library("effects")
knitr::opts_chunk$set(fig.width = 6, fig.height = 4.5, fig.align = "center")
```


## Multiple Regression

[**Let's take a look at the iris data set!**]{.medium}

```{r, echo=FALSE}
#loading data

iris<-datasets::iris

head(iris)

summary(iris)
mod1 <- lm(Petal.Length~Sepal.Length*Sepal.Width,data= iris)

```

:::puzzle

**You Try!**

- Build a multiple regression model with `Petal Length` as the outcome and `Sepal Length` and `Sepal Width` as predictors.
- Name it `mod_cont`.
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app1/",
                    id = "learnr-30-app1")

```

Let’s interpret those coefficients!

- What does the estimate for the intercept mean?
- What do the coefficients for each variable mean?
- If the Estimate for Sepal Length was 2 and the Standard Error was 0.5, what would the *t*-value be?
- What do the *p*-values mean?
- What does the $R^2$ stand for?

:::puzzle

**You try!**

- Let's build a multiple regression model with Petal Length as the outcome and Sepal Length and Species as predictors. 
- Name it `mod_cat`.
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app2/",
                    id = "learnr-30-app2")

```

Let’s interpret those coefficients!

- What is the reference (base) category?
- What does the estimate for the intercept mean?
- What do the coefficients for each variable mean?


## Continuous Predictors Interacting

:::puzzle

**You try!**

- Let's build a regression model with Petal Length as the outcome and Sepal Length and Sepal Width as predictors that interact.
- Name it `mod1`.
- We'll reference `mod1`throughout the assignment, so don't forget what it stands for!
:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app3/",
                    id = "learnr-30-app3")

```

Let’s interpret those coefficients!

- What does the estimate for the intercept mean?
- What do the coefficients for each variable mean?
- What happened to the coefficient of Sepal Length compared to the multiple regression model `mod_cont`?


:::quiz
**Quick Quiz**
:::
```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-quiz1/",
                    id = "learnr-30-quiz1")

```


`r emo::ji("index_pointing_up")` Good job! You can work with continuous variable interactions!


## Simple Slopes

**A great way of understanding interactions is by visualizing them!**

Example 1:

First, choose which package you want to use. We'll start with `sjPlot`.

```{r example1}
library(sjPlot)
```

Plug in your regression model(we'll use `mod1`) and plot the interaction!

```{r example2}
plot_model(mod1, type = "pred",terms = c("Sepal.Width","Sepal.Length"))
```

Now, we'll look at these interactions for `mod1` when `Sepal.Width` equals 2 and 3. 

```{r}

plot_model(mod1, type = "pred", terms = c("Sepal.Width", "Sepal.Length[2,3]"))

```

:::puzzle
**You try!**

- Let's visualize this model: `Sepal.Width ~ Sepal.Length*Petal.Length` using the `sjPlot` package
- Name your regression model `mod_width`
- Then let's visualize this models when `Sepal.Length` is 4 and 6

:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app4/",
                    id = "learnr-30-app4")

```

Let's look at these interactions across `Species`. For this, we'll need to create a new regression model first.

```{r}

mod1_species<-lm(Petal.Length~Sepal.Length*Sepal.Width+Species, data=iris)

plot_model(mod1_species, type = "pred", terms = c("Sepal.Length", "Sepal.Width", "Species"))

```


Example 2:

There are many ways of exploring an interaction. Now we're going to try using the `reghelper` package to look at the significance of our simple slopes at specific levels.

Using `mod1`, let's see if Sepal Width is a significant predictor of Petal Length at different levels of Sepal Length (specifically when Sepal Length is 5 and 6).

```{r}

simple_slopes(mod1, levels = list(Sepal.Length = c(5,6)))

# what do these results mean?
```


:::puzzle
**You try!**

- Now let's see if Sepal Length is a significant predictor of Sepal Width at different levels of Petal length (specifically when Petal Length is 4 and 6).
- Please use the `reghelper` package
- HINT: check slide 53

:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app5/",
                    id = "learnr-30-app5")

```

Example 3:

Just like how there are many ways of exploring simple slopes, there are also many ways of visualizing interactions!

Let's visualize `mod1` using the package `ggeffects`. 

```{r}

predict<- ggpredict(mod1, terms =
                      c("Sepal.Width","Sepal.Length"))

plot(predict)

```

There's something cool about this function, though. Check out the data that `ggpredict` creates!

```{r}
predict
```

What does this output mean?


:::puzzle

**Make it pretty!**

- Let's visualize `mod_width` using the same package!
- This time, set color palette to "eight"
- Here's a nice website for reference https://strengejacke.github.io/ggeffects/reference/plot.html

:::

```{r, echo=FALSE}

blogdown::shortcode("learnr",
                    url = "https://shelly-cooper.shinyapps.io/30-learnr-app6/",
                    id = "learnr-30-app6")

```



What is this graph telling you?


<small>Massive shout out to the Fall 2022 AI Selin Toprakkiran for creating this excellent Practice Set!</small>